{\bf \Large THESISSSS}

### 2023-05-23 draft.tex
general introduction done

\author{Emil Mirzayev}
%\author[2,3]{Zakaria Babutsidze}
%\affil[1]{University College London}
%\affil[2]{SKEMA Business School}
%\affil[3]{Sciences Po Paris}
%\date{}

@@

    This thesis demonstrates the existence of context effects in multi-attribute, multidimensional settings and develops methodologies on enhancing recommender systems design with context effects. Also, this study investigates context arising outside the choice set, namely, the tools that allows users to express their preferences.

@@

The context in which a consumer makes decisions significantly influences their choices. This thesis investigates the role of context effects in consumer decision-making and applies this knowledge to enhance recommender systems in online marketplaces. Leveraging a computational model known as MDbS, we examine the impact of context effects in real-world choice data. Our findings indicate that these effects, previously studied primarily in experimental settings, also manifest in field scenarios. This insight enables online platforms to utilize mathematical decision-making models to better tailor their product lists and recommendations.

Our work extends beyond current measures of context effects by making contextual choice modeling applicable to a wider range of observational data. This approach paves the way to study context effects in various settings and opens up opportunities for optimizing recommender systems in e-commerce platforms. We also propose a novel approach to tackle the user-side continuous cold-start problem in recommender system design, using the context information of the product menu to generate relevant choice sets.

Further, we delve into the role of user control on recommender systems' acceptance, using the Technology Acceptance Model (TAM) as a theoretical framework. Our findings suggest that users perceive easy-to-use recommender systems as more useful and intend to use them more. We also find that different control mechanisms can have varying impacts on user experience, leading to diverse effects on user acceptance of recommender systems.

This thesis offers valuable insights for online marketplaces and recommender systems, showcasing the benefits of considering context effects in product presentation and user decision-making processes. Our research also identifies future directions for enhancing user control in recommender systems, including the integration of large language models for more intuitive and transparent user interactions. By recognizing the dynamic nature of user preferences and the significant role of context effects in decision-making, our work contributes to the design of more effective and user-centric recommender systems.

TWO TYPE OF CONTEXT EFFECTS - THE ONES WITHIN THE CHOICE SETS AND ONES WITHOUT. ONES WITHOUT ARE GIVEN BY RECOMMENDER SYSTEMS. ONE WITHIN COMES WITHIN. MY ULTIMATE GOAL IS TO CONTRIBUTE TO CONTEXT EFFECT APPLICATIONS TO RECOMMENDER SYSTEMS.

FIRST I WANT TO CONTRIBUTE TO UNDERSTANDING OF CONTEXT EFFECTS AND THEIR DETECTION IN MULTIDIMENSIONAL MULTIATTRIBUTE CHOICE SETS -> THIS IS THE FIRST PAPER, SIMULATION USING MATHEMATICAL MODEL TO ACCOUNT FOR CONTEXT EFFECTS.
THEN I WANT TO CONTRIBUTE TO CONTEXT EFFECT DIFFERENTLY, BY ACCOUNTING FOR THEM USING CLUSTERING.
THEN I WANT TO CONTRIBUTE TO CONTEXT EFFECT BY DISENTANGLING BIG THREE CONTEXTS

ONE OF THE MAIN GOALS IS TO COME UP WITH ABILITY OF EXTENDING CONTEXT EFFECT APPLICATIONS TO RECOMMENDER SYSTEMS
ANOTHER GOAL IS TO UNDERSTAND THE CONTEXTUAL EFFECTS ARISING FROM MULTIDIMENSIONAL CHOICE SETS
ALSO, TO UNDERSTAND HOW

\subsection{Data and research methodology}

\subsection{Placeholder}

BLABLABLA

@@

\subsubsection{Observational Data}\label{section:observationalDataDescription}

@@

\subsubsection{Experimental data}

@@

\subsubsection{Parameter optimization}

@@

\subsubsection{Observational data results}

@@

\subsubsection{Experimental data results}

@@

\section{Choice modeling with context effects:\\
Generalization to multi-option and multi-attribute settings
}  \footnote{This chapter is based on a joint work with my supervisor Zakaria Babutsidze and also William Rand, Nobuyuki Hanaki, Ismael Rafai, Rodrigo Acuna Agost and Thierry Delahaye}

@@

HERE SOME BRIEF INTRO TO REFRESH CONTEXT EFFECTS

Over the years, multiple empirical models have been developed to model context effects. Empirical approaches usually model context effects either in the structural part of utility or in the error covariance part \citep{kamakuraSrivastava84, dotsonEtAl18}. Some of these models have the capacity to account for multiple effects at the same time \citep{tverskySimonson93, orhun09}. These models extend a classical random utility model \citep{mcfadden01} in multiple directions using discrete choice modeling \citep{benAkivaLerman85}. However, Rooderkerk, Van Heerde, and Bijmolt \citeyearonly{roodrkerkEtAl11} present a unifying model taking into account all three context effects. Instead of using advanced statistical techniques for remedying violations of utility maximization assumptions associated with the existence of context effects \citep{luce59}, their approach hinges on additive specification and ex ante calculation of individual measures for each of the three context effects for each item in the menu. Namely, authors assume that choice estimator is additive in three context effects (along with a generic preference-driven part) and develop the methodology of quantifying three effects for each alternative prior to calculating the estimator. This is a particularly flexible approach which also ensures that researcher does not run into computational difficulties (i.e., $NP$-hard calculations). We follow the suite and formulate the utility that a consumer c attaches to an option i, under a given menu m, as being additive in two parts:

@@

\subsection{Generalizing context effect measures}
\subsubsection{Approach to generalization}

@@

\subsubsection{Attraction effect}

@@

\subsubsection{Compromise effect}

@@

 \subsubsection{Similarity effect}

@@

\subsubsection{Observational data}\label{section:additionalPreprocessingObservationalData}

@@

Recommender systems (RS) have been an important part of our daily lives thanks to the rise of the Internet. RS are software tools and/or algorithms which match users to items \citep{mahmood2009improving} \footnote{Although there are other tools which create suggestions for users (such as Interactive Decision Aids, Recommender agents and etc.), for the purposes of this study I refer to recommender systems and bound the definition using the one by Mahmood \citeyearonly{mahmood2009improving}. These other tools have some theoretical differences which makes orthogonal to the purposes of this study. However, I discuss, differentiate between them and define the boundary conditions in the chapter \ref{chapter:UserControlAndRS}}. One example is Netflix, which recommends a movie similar to the one the user just watched. The general purpose of any RS is to help users who do not have sufficient knowledge or experience, or the capacity to evaluate the item pool fully.

@@

Recently, some progress was made in solving the user-side cold-start problem after the introduction of contextual information into recommender systems. As a result of this effort, Context Aware Recommender systems were introduced \citep{adomavicius2011context}. In this approach the context refers to the time of the choice, the location or socio-demographic characteristics of the decision-maker etc.  Some approaches have been very successful by combining contextual information with collaborative filtering \citep{aharon2013off,bykau2013coping,saveski2014item}. Utilizing baseline information for new users \citep{kluver2014evaluating} and using social network data \citep{guy2009personalized} have also been proven to overcome the cold-start problem to a certain extent.

@@

It was previously observed that users' preferences are prone to change over time. They become bored with the similarity of the items they have consumed and suggested and want more personalization \citep{songWhenHowDiversify2019}. This makes RS adaptation to dynamic user preference extremely important. To mitigate RS low ability to "evolve" alongisde with the users' preferences, system engineers have introduced user control mechanisms to allow users to participate in personalization.

@@

\subsubsection{Recommender systems}

@@

\subsection{User control}

@@

\subsubsection{Hypothesis development}

@@

>>>>

\author{Emil Mirzayev}

@@

    This thesis demonstrates the existence of context effects in multi-attribute, multidimensional settings and develops methodologies on enhancing recommender systems design with context effects. Also, this study investigates context arising outside the choice set, namely, the tools that allows users to express their preferences and how to make recommender systems better through them.

@@

The main objective of this thesis is to make contribution towards understanding the choice context and applying it to improve recommender systems. Thesis makes effort by identifying context effects arising from the choice set and developing a methodology to implement this information in recommender systems design. Furthermore, it analyzes the effectiveness of user control mechanisms, which allow individuals to inform the system about their preferences and also amend them.

Recommender systems have become a crucial ally in the vast landscape of digital world. They help individuals to navigate through the plethora of alternatives and find what they are looking for. They achieve this by using sophisticated algorithms that analyze the wide space of items, users and their interactions with each other. System designers and businesses thrive to maximize the accuracy of recommendations, meaning they want to maximize the consumption of the recommendations. However, recent research stream shows that, accuracy is not the only way to make recommender systems effective. Other factors like diversity, serendipity \citep{kaminskas2016diversity}. Moreover, the black-box nature of the algorithms do not allow to understand the reasons behind a particular user's behaviour \citep{samih2021exmrec2vec, kotkovSurveySerendipityRecommender2016}. Hence, more and more interest is driven towards the understanding and utilizing the choice context \citep{adomavicius2005toward}. Choice context has received interest in scenarios where user or item information is scarce \citep{scheinMethodsMetricsColdStart}.

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates and uses two of them. The first one posits that choice context \footnote{For the remainder of the paper, I will follow the previous literature \citep{truebloodEtAl13} and refer this definition of choice context as context effects} is "the availability and the nature of choice alternatives" \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Context also referred to as "the time and content of the choice, the location or socio-demographic characteristics of the decision-maker..." \citep{adomavicius2011context}.

However, modeling for these effects mathematically was a challenge because the existing models suffered from independence of irrelevant alternatives \citep{luce59}, meaning these models treated each alternative in isolation. This has shifted the scholar's interest towards computational models \citep{usher2001time, roe2001multialternative, trueblood2014multiattribute, noguchi2018multialternative}. Yet, none of these models have been to real-world choice data and as a result, their applicability to field data remained an open question. I address this gap in the first study where I apply a computational model to a field data of high heterogeneity among dimensions. Building on my results in my second study I delve into the choice modeling more and develop a methodology to account for three main context effects studied in the literature \citep{truebloodEtAl13}.

When new users or items are introduced to recommender systems, they fail to function as intended because of the lack of information their algorithms need. In such cases, even when they do provide recommendations, they are far from personalized \citep{lika2014facing}. It is considered to be a key challenge in recommender system design \citep{park2009pairwise}. Previous literature has applied various methods to overcome this problem \citep{aharon2013off, bykau2013coping, saveski2014item, guy2009personalized}. However, the limitations of these approaches are that they ignore the context effects and concentrate on scenarios where information scarcity is temporary. Using the findings from my second study, decision-making and choice modeling literature I propose an innovative solution to address continious cold-start issue in recommender system design by utilizing menu contextual information to generate relevant choice sets through a two-step choice modeling method.

Research agrees on importance of metrics beyond accuracy for recommender systems \citep{kaminskas2016diversity}. It is in the best interest of online marketplaces to provide users with recommendations which are not only in the category "exactly what I want", but also "I never thought I would have liked this one" \citep{kotkovSurveySerendipityRecommender2016}. This can not only boost sales \citep{songWhenHowDiversify2019}, but also increase the satisfaction of users \citep{knijnenburgExplainingUserExperience2012}. Albeit, it was observed that the preferences of individuals are not stable and tend to change and system designers have proposed various tools to enable users signal their preference shifts \citep{bostandjiev2012tasteweights,hijikata2012relation}. However, to better understand it is necessary to study those tools in isolation from the context arising from the choice set and concentrate purely on context arising from the user side, e.g their preferences \citep{adomavicius2011context}. In my last study I address this gap by conducting an online experiment and applying Technology Acceptance Model \citep{davis1985technology} to measure users' acceptance of such tools.

All in all, all four studies aim to contribute to our understanding of context and provide applications of recommender system design using this information.

@@

\textbf{Observational Data}\label{section:observationalDataDescription}

@@

\textbf{Experimental data}

@@

\textbf{Parameter optimization}

@@

\textbf{Observational data results}

@@

\textbf{Experimental data results}

@@

\section{Choice modeling with context effects:
Generalization to multi-option and multi-attribute settings
}  \footnote{This chapter is based on a joint work with my supervisor Zakaria Babutsidze and also William Rand, Nobuyuki Hanaki, Ismael Rafai, Rodrigo Acuna Agost and Thierry Delahaye}

@@

Over the years, multiple empirical models have been developed to model context effects. Empirical approaches usually model context effects either in the structural part of utility or in the error covariance part \citep{kamakuraSrivastava84, dotsonEtAl18}. Some of these models have the capacity to account for multiple effects at the same time \citep{tverskySimonson93, orhun09}. These models extend a classical random utility model \citep{mcfadden01} in multiple directions using discrete choice modeling \citep{benAkivaLerman85}. However, Rooderkerk, Van Heerde, and Bijmolt \citeyearonly{roodrkerkEtAl11} present a unifying model taking into account all three context effects. Instead of using advanced statistical techniques for remedying violations of utility maximization assumptions associated with the existence of context effects \citep{luce59}, their approach hinges on additive specification and ex ante calculation of individual measures for each of the three context effects for each item in the menu. Namely, authors assume that choice estimator is additive in three context effects (along with a generic preference-driven part) and develop the methodology of quantifying three effects for each alternative prior to calculating the estimator. This is a particularly flexible approach which also ensures that researcher does not run into computational difficulties (i.e., $NP$-hard calculations). I follow the suite and formulate the utility that a consumer c attaches to an option $i$, under a given menu $m$, as being additive in two parts:

@@

\subsection{Generalizing context effect measures}
\textbf{Approach to generalization}

@@

\textbf{Attraction effect}

@@

\textbf{Compromise effect}

@@

 \textbf{Similarity effect}

@@

\textbf{Observational data}\label{section:additionalPreprocessingObservationalData}

@@

Recommender systems (RS) have been an important part of our daily lives thanks to the rise of the Internet. RS are software tools and/or algorithms which match users to items \citep{mahmood2009improving} \footnote{Although there are other tools which create suggestions for users (such as Interactive Decision Aids, Recommender agents and etc.), for the purposes of this study I refer to recommender systems and bound the definition using the one by Mahmood \citeyearonly{mahmood2009improving}. These other tools have some theoretical and implementation differences which makes them orthogonal to the purposes of this study. However, I discuss, differentiate between them and define the boundary conditions in the chapter \ref{chapter:recommenderSystemsAgentsDifferentiation}}. One example is Netflix, which recommends a movie similar to the one the user just watched. The general purpose of any RS is to help users who do not have sufficient knowledge or experience, or the capacity to evaluate the item pool fully.

@@

Recently, some progress was made in solving the user-side cold-start problem after the introduction of contextual information into recommender systems. As a result of this effort, Context Aware Recommender systems were introduced \citep{adomavicius2011context}. In this approach the context refers to the time and content of the choice, the location or socio-demographic characteristics of the decision-maker etc.  Some approaches have been very successful by combining contextual information with collaborative filtering \citep{aharon2013off,bykau2013coping,saveski2014item}. Utilizing baseline information for new users \citep{kluver2014evaluating} and using social network data \citep{guy2009personalized} have also been proven to overcome the cold-start problem to a certain extent.

@@

It was previously observed that users' preferences are prone to change over time. They become bored with the similarity of the items they have consumed and suggested and want more personalisation \citep{songWhenHowDiversify2019}. This makes RS adaptation to dynamic user preference extremely important. To mitigate RS low ability to "evolve" alongside with the users' preferences, system engineers have introduced user control mechanisms to allow users to participate in personalisation.

@@

\textbf{Recommender systems}\label{chapter:recommenderSystemsAgentsDifferentiation}

@@

\textbf{User control}

@@

\textbf{Hypothesis development}

@@

\section{General conclusion}

In this

 I show that not only this model is able to capture context succesfully, but also information it provides has meaningful contribution to traditional choice model. One can utilize this information to partly address the drawbacks of such models like logit and probit. However, the nature of computational models only allows them to account for the context as aggregate. To address this drawback in

 This research introduced a flexible framework for accounting for context effects across various settings and with large and complex product menus. The methodology proved to adaptable to inconsistent menu sizes and is not confined to a single empirical application. Its potential extends to large-scale data from e-commerce websites and offline sources, enabling context effect research without costly experimental interventions. Consequently, this method can aid in designing optimal offers and enhancing recommender systems, especially for online outlets presenting large arrays of options.

 The approach remains robust to selection criteria and contextualization method used, highlighting the efficiency and speed of implementation which is of essence in online environments. It further underscores the critical role of choice set context effects in influencing user decisions and demonstrates the potential of integrating these effects into recommender system design.
